---
title: "Applying CMAP color schemes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applying CMAP color schemes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 400/670,
  fig.retina = 4,
  fig.align = "center",
  dev = "ragg_png"
)

library(tidyverse)
library(cmapplot)
```

There are `r nrow(get_cmapplot_global("palettes"))` color palettes in the cmapplot package, across three categories: discrete, sequential, and divergent. Mirroring the underlying functionality of ggplot2, cmapplot can apply these color palettes as either `discrete` or `continuous` scales to either the `color` (outline) or `fill` attributes of a ggplot.

## Using CMAP colors in plots

### Discrete scales

Discrete color scales are used to assign colors to discrete bins, such as the lines in a time series line chart. Discrete scale functions work with any palette -- discrete, sequential, or divergent. Add a discrete color scale by calling either the `cmap_fill_discrete()` or `cmap_color_discrete()` function in your plot construction. 

Note that discrete scales will automatically interpolate additional colors if the dataset has more colors than the underlying palette. This can be helpful but is not ideal for finished graphics.

```{r color_discrete, message = FALSE}
df <- dplyr::filter(grp_over_time, category == "Goods-Producing")

ggplot(data = df) +
  geom_line(mapping = aes(x = year, y = realgrp, color = cluster), 
            size = 1.25) +
  scale_x_continuous(breaks = seq(from = 2007, to = 2017, by = 2)) +
  cmap_color_discrete(palette = "community", reverse = TRUE) +
  theme_cmap() +
  ggtitle("Real GRP of goods-producing clusters over time")
```

### Race/ethnicity scale

In order to maintain a consistent data language, CMAP uses specific colors for displaying data based on race and ethnicity. To map these colors to each demographic group, use `cmap_fill_race()` or `cmap_color_race()`. 

In the arguments, specify the case-sensitive name of each group as it appears in your data. The functions can be used even if your dataset does not contain every race/ethnicity category — simply omit the parameters for the missing categories.

```{r color_race, message = FALSE}
df <- dplyr::filter(
  traded_emp_by_race,
  Race != "Regional average" & variable == "SpecializedTraded"
)

ggplot(data = df) +
  geom_col(mapping = aes(x = Race, y = value, fill = Race)) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  cmap_fill_race(white = "White",
                 black = "Black",
                 hispanic = "Hispanic",
                 asian = "Asian",
                 other = "Other") +
  theme_cmap(show.legend = FALSE) +  # Legend is redundant in this example
  ggtitle("Percent employed in specialized traded clusters, by race")
```

### Continuous scales
Continuous color scales work with sequential and divergent palettes. Add a continuous scale by calling either the `cmap_fill_continuous()` or `cmap_color_continuous()` function in your plot construction. If you're using a divergent palette, you can specify the midpoint where the divergence begins (default is zero).

```{r color_continuous, message = FALSE}
df <- dplyr::filter(
  percentile_wages,
  cluster %in% c("Biopharmaceuticals", "Hospitality and Tourism", "Paper and Packaging")
)

ggplot(data = df) +
  geom_point(mapping = aes(x = cluster, y = wage, color = percentile),
             size = 5) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_discrete(labels = scales::label_wrap(18)) +
  coord_flip() +
  cmap_color_continuous(palette = "red_teal", middle = 50) +
  theme_cmap() +
  ggtitle("Wage percentiles for key clusters")
```

### Highlighting categories

If you want to draw attention to a specific group in your graph, use `cmap_fill_highlight()` or `cmap_color_highlight()`. This will make your highlighted group one color and all other groups identical in another color. Specify the vector in your data that determines the groups, and then the value of the group to be singled out. Note that this *must* be the same vector specified in the `fill`/`color` aesthetic. The highlight and non-highlight colors have defaults, but can be changed with the `color_value` and `color_other` parameters, respectively. 

More than one category can be highlighted as well, by feeding a vector of values into the the `value` argument. See `?cmap_fill_highlight` for details.

```{r color_highlight, message = FALSE}
df <- dplyr::filter(transit_ridership, year == 2019)

ggplot(data = df) +
  geom_col(mapping = aes(x = system, y = ridership, fill = system)) +
  cmap_fill_highlight(field = transit_ridership$system,
                      value = "metra") +
  theme_cmap(show.legend = FALSE) +  # Legend is redundant in this example
  ggtitle("Annual passenger trips by service, in millions")
```


## Available color palettes
Palettes are stored in a tibble in the `cmapplot_globals` environment. This tibble can be accessed with `get_cmapplot_global("palettes")`, with specific details about any specific palette most easily accessed with `fetch_pal()`. Palettes can be visualized in the plot window with `viz_palette()` and `viz_gradient()`. This table lists all palettes currently available in the package:

```{r palettes, echo = FALSE, results = 'asis'}

pals <- get_cmapplot_global("palettes")
discrete <- filter(pals, type == "discrete")[["name"]]
sequential1 <- filter(pals, type == "sequential", !stringr::str_detect(name, "_"))[["name"]]
sequential2 <- filter(pals, type == "sequential", stringr::str_detect(name, "_"))[["name"]]
divergent <- filter(pals, type == "divergent")[["name"]]

maxlength <- max(length(discrete), length(sequential1), length(sequential2), length(divergent))
length(discrete) <- length(sequential1) <- length(sequential2) <- length(divergent) <- maxlength

out <- tibble(Discrete = discrete, 
              Sequential = sequential1,
              `Sequential (multi-hue)` = sequential2,
              Divergent = divergent)
out[is.na(out)] <- ""

knitr::kable(out)
```

### Discrete palettes
Use these palettes with `cmap_fill_discrete()` or `cmap_color_discrete()`:
```{r show_discrete, echo = FALSE, fig.asp = 300/670}
nms <- dplyr::filter(get_cmapplot_global("palettes"),  type == "discrete", name != "race")$name
purrr::walk(nms, viz_palette)
```

The race palette is a special discrete palette, which should be used with the specific functions `cmap_fill_race()` or `cmap_color_race()`:
```{r show_discrete_palettes, echo = FALSE, fig.asp = 300/670}
viz_palette("race")
```

### Sequential palettes
Use these palettes with `cmap_fill_continuous()`, `cmap_fill_discrete()`, or their `color` variants.
```{r show_sequential, echo = FALSE, fig.asp = 300/670}
nms <- dplyr::filter(get_cmapplot_global("palettes"),  type == "sequential")$name
purrr::walk(nms, viz_gradient)
```

### Divergent palettes
Use these palettes with `cmap_fill_continuous()` `cmap_fill_discrete()`, or their `color` variants.
```{r show_diverent, echo = FALSE, fig.asp = 300/670}
nms <- dplyr::filter(get_cmapplot_global("palettes"),  type == "divergent")$name
purrr::walk(nms, viz_gradient)
```

